Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-19 paper/manuscript.tex
revisions to Discussion

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 on viral growth in cell culture, and demonstrated that these measurements can be informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, strains on the trunk of the phylogenetic tree tend to have mutations that are beneficial for viral growth in our experiments when compared to side-branch mutations.
Differentiating between trunk and side branch lineages can enable forecasting of influenza virus evolution by identifying which strains are likely to succeed in nature.
The fact that our experiments are able to reveal differences between trunk and side branch lineages highlight the potential for incorporating experimental measurements of the non-antigenic effects of mutations for evolutionary forecasting.

Surprisingly, our simple experimental measurements of viral growth in cell culture helped to elucidate differences in fitness of viral strains in nature.
Mutations may have effects in the laboratory that are distinct from their effects in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}, and our experiments certainly don't capture all of the relevant selective pressures on HA that exist in nature.
Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
Yet, prior to our work presented here, there were no comprehensive studies of the functional effects of mutations to HA on viral fitness outside of antigenicity.
Previous quantitative modeling work either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or made the simplifying assumption that all mutations outside epitope sites had equivalent deleterious effects on viral fitness~\cite{luksza2014predictive}.
While our measurements may be imperfect, we have demonstrated that these measurements are still of utility in distinguishing the selective advantage of successful influenza viral lineages over $\sim$50 years of evolution.
This also suggests that measurements in experimental models that more faithfully recapitulate the selective forces in nature should be of even greater utility for differentiating evolutionarily successful viral lineages.

In addition, we are measuring the effects of mutations in a single HA variant from the Perth/2009 strain and extrapolating these measurements to other seasonal human H3N2 strains across a 50-year timespan.
However, these extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
% and the extent to which this occurs is an important question in its own right.
The observation that our measurements score trunk mutations more favorably than side branch mutations across most of H3N2 evolution suggests that for relatively closely related strains, the mutational effects are not shifted to a large extent.
On the other hand, when we compared our measurements for H3 to prior measurements on H1, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
Further analysis of how mutational effects shift as protein variants diverge can provide insight into how broadly the results of a given experiment can be generalized across variants.

Importantly, the antigenic properties of HA are integral in determining strain success, but we did not study the effect of mutations on antigenicity.
However, our high-throughput experimental approach can be harnessed to profile mutations that confer escape from antibodies, as has been performed for the WSN/1933 H1 HA~\cite{doud2017complete,doud2017quantifying} and HIV Env~\cite{dingens2017comprehensive}.
In future, these antigenic studies can be combined with functional measurements to inform forecasting of the evolution of seasonal influenza virus.

@@

\begin{suppdata}
\caption{\label{suppdata:rep1_prefs}
CSV file providing the amino-acid preferences for replicate 1 in sequential HA numbering.
}
\end{suppdata}

\begin{suppdata}
\caption{\label{suppdata:rep2_prefs}
CSV file providing the amino-acid preferences for replicate 2 in sequential HA numbering.
}
\end{suppdata}

\begin{suppdata}
\caption{\label{suppdata:rep3-1_prefs}
CSV file providing the amino-acid preferences for replicate 3-1 in sequential HA numbering.
}
\end{suppdata}

\begin{suppdata}
\caption{\label{suppdata:rep3-2_prefs}
CSV file providing the amino-acid preferences for replicate 3-2 in sequential HA numbering.
}
\end{suppdata}

\begin{suppdata}
\caption{\label{suppdata:avg_prefs}
CSV file providing the across-replicate averaged and re-scaled amino-acid preferences in H3 numbering.
}
\end{suppdata}

>>>>

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture, and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, Luksza and Lassig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
Our work suggests that deep mutational scanning can provide such data.

It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
Indeed, a vast amount of work in virology has chronicled the many ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
Mutations in viral genes other than HA are also certainly important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
Given these caveats, it might seem surprising that simply measuring viral growth in cell culture can be informative about the success of viral strains in nature.
Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or simply assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and represent a vast improvement over no information at all.
That this is true is demonstrated by the fact that these measurements have utility for distinguishing successful and unsuccessful viral lineages over $\sim$50 years of H3N2 evolution.
We suspect that performing the experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures~\cite{?}) might further improve their utility.

We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
The fact that our measurements score trunk mutations more favorably than side branch ones across a 50-year timespan suggests that the mutational effects have not shifted to dramatically among these relatively closely related sequences.
On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.

Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{?}.
Indeed, some of our results (e.g., Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
But so far, experiments using this approach have either failed to examine antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.

@@

\begin{suppdata}
\caption{\label{suppdata:rep1_prefs}
Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.
The unscaled preferences for replicates 1, 2, 3-1, and 3-2 are each in a separate tab of the file.
Additional tabs give the across-replicate averaged and re-scaled amino-acid preferences in sequential numbering and in H3 numbering as shown in Figure~\ref{fig:logoplot}.
There are also tabs that give the conversion from sequential to H3 numbering, and the epitope sites in H3 numbering as defined by Wolf et al and used in this study.
Each tab can simply be exported to CSV for computational analyses.
}
\end{suppdata}
